---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# eqtlgentools

<!-- badges: start -->
<!-- badges: end -->

The goal of eqtlgentools is to generate beta and standard error values for the [cis-eQTL](https://molgenis26.gcc.rug.nl/downloads/eqtlgen/cis-eqtl/README_cis) dataset from the [eQTLGen Consortium](https://www.eqtlgen.org/). This requires [allele frequencies](https://molgenis26.gcc.rug.nl/downloads/eqtlgen/cis-eqtl/README_allele_frequency) from a separate file. Files are publicly available from [here](https://www.eqtlgen.org/cis-eqtls.html)

The calculations for converting Z-scores to betas and standard errors are documented in this [here](https://molgenis26.gcc.rug.nl/downloads/eqtlgen/cis-eqtl/README_allele_frequency).

## Installation

Install from [github](https://github.com/rmgpanw/eqtlgentools) with:

```{r eval=FALSE}
devtools::install_github("rmgpanw/eqtlgentools")
```

## Usage

Run the following code (replacing the file paths with your own) in R to generate a new file with beta and standard error columns appended to the main cis-eQTL file. Adjust the `chunk_size` parameter if needed.

```{r eval=FALSE}
append_beta_se_to_eqtlgen_cis_eqtl(
  eqtlgen_cis_eqtl_filepath = "PATH/TO/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt",
  eqtlgen_cis_eqtl_maf_filepath = "PATH/TO/2018-07-18_SNP_AF_for_AlleleB_combined_allele_counts_and_MAF_pos_added.txt",
  outfile_path = "PATH/TO/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded-beta-se.txt",
  chunk_size = 1000000
)
```

For a full list of available functions, please see [`Reference`](https://rmgpanw.github.io/eqtlgentools/reference/index.html) section on [package website](https://rmgpanw.github.io/eqtlgentools/).
