---
title: "Explore full eQTLGen cis-eQTL dataset"
output: html_notebook
---

```{r include=FALSE}
library(data.table)
library(tidyverse)
```

```{r include=FALSE}
# cis-eqtl data: read first 100 rows
eqtl_100rows <- fread(file = "../tests/data/raw/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt",
  nrows = 100)

# cis-eqtl data: read all rows, SNP col only
eqtl_SNP_col <- fread(file = "../tests/data/raw/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt",
    select = c("SNP"))

# allele frequency data
allele_freq <- fread(file = "../tests/data/raw/2018-07-18_SNP_AF_for_AlleleB_combined_allele_counts_and_MAF_pos_added.txt",
                     select = c("AlleleB_all", "SNP"))

# source scripts
source("../R/beta_se_from_zscore_maf.R")
source("../R/process_file_chunked.R")
```

# Overview

> This is a throw away script. Explores the eQTLgen cis-eqtl dataset (all functions now in `R` directory of package). Final code chunk is for appending beta and SE columns to the raw data, which resides in `tests/data`.

**Aim:** to convert z-scores to betas and SEs in the [cis-eQTL dataset](https://www.eqtlgen.org/cis-eqtls.html) from [eQTLGen](https://www.eqtlgen.org/) ([click here to donwnload dataset](https://molgenis26.gcc.rug.nl/downloads/eqtlgen/cis-eqtl/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt.gz)).

- Convert z-scores to betas/SEs: [see README](https://molgenis26.gcc.rug.nl/downloads/eqtlgen/cis-eqtl/SMR_formatted/README_cis_full_SMR). ***Note: there is a missing closing bracket at the end of both formulae. I double checked the formulae with this [biostars post](https://www.biostars.org/p/235966/).
- Allele frequencies (needed for conversion):

  - [Download link](https://molgenis26.gcc.rug.nl/downloads/eqtlgen/cis-eqtl/2018-07-18_SNP_AF_for_AlleleB_combined_allele_counts_and_MAF_pos_added.txt.gz)
  - [README](https://molgenis26.gcc.rug.nl/downloads/eqtlgen/cis-eqtl/README_allele_frequency)

# Allele frequencies

- All values in SNP col are unique
- No `NA` values in SNP col

```{r}
nrow(allele_freq)

length(unique(allele_freq$SNP))

sum(is.na(allele_freq$SNP))
```

# eqtl file

- All values in SNP col are unique
- No `NA` values in SNP col

```{r}
nrow(eqtl_SNP_col)

length(unique(eqtl_SNP_col$SNP))

sum(is.na(eqtl_SNP_col$SNP))
```

# Convert Z-scores to beta and SE

beta = Z-score / sqrt(2 * alleleFreq * (1-alleleFreq) * (N + Z-score^2)
se = 1 / sqrt(2 * alleleFreq * (1-alleleFreq) * (N + Z-score^2)

## Merge eQTL data with allele frequencies

According to README files, the `AssessedAllele` in the eQTL dataset corresponds to `AlleleB` in the allele frequency file. Therefore, the allele frequency to use is in column `AlleleB_all`.

```{r}
eqtl_100rows_maf <- eqtl_100rows %>% 
  left_join(allele_freq, by = "SNP")

# raise error if these columns do not correspond
if(!(all(eqtl_100rows_maf$AssessedAllele == eqtl_100rows_maf$AlleleB))) {
  stop("Error! Effect allele column differs between cis-eQTL and maf files.")
}
```

## Mutate betas and SEs

```{r}
# get allele frequencies
eqtl_100rows_maf <- eqtl_100rows_maf %>% 
  mutate(
    beta = Zscore / sqrt(2 * AlleleB_all * (1-AlleleB_all) * (NrSamples + Zscore^2)),
    se = 1 / sqrt(2 * AlleleB_all * (1-AlleleB_all) * (NrSamples + Zscore^2))
  )
```

# Repeat using functions

## Test informally

```{r}
# mutate beta and se in new df
test <- eqtl_100rows %>%
  left_join(allele_freq, by = "SNP")

test$beta <- beta_se_from_zscore(
  df = test,
  zscore = "Zscore",
  maf = "AlleleB_all",
  sample_n = "NrSamples",
  beta_se = "beta"
)

test$se <- beta_se_from_zscore(
  df = test,
  zscore = "Zscore",
  maf = "AlleleB_all",
  sample_n = "NrSamples",
  beta_se = "se"
)

# check that these equal results above - error if not
if ((!all(test$beta == eqtl_100rows_maf$beta)) |
    (!all(test$se == eqtl_100rows_maf$se))) {
  stop("Different results produced for beta/se")
}
```


```{r}
# function specific to this data
test <- mutate_beta_se_from_zscore_eqtlgen_cis_eqtl(eqtl_100rows, 
                                                    allele_freq)

# check that these equal results above - error if not
if ((!all(test$beta == eqtl_100rows_maf$beta)) |
    (!all(test$se == eqtl_100rows_maf$se))) {
  stop("Different results produced for beta/se")
}
```

## Test using `testthat`

See script under `tests/testthat` directory for tests

# Process file in chunks

I created a file called `test.txt` which is the first 1000 rows of the eQTL data. Aim here is to write a new file with additional beta and se columns.

```{bash}
head -n 1000 ../data/raw/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt > ../tests/data/raw/test.txt
```


```{r}
# write new file with additional beta/SE cols
process_file_chunked(
  infile_path = "../tests/data/raw/test.txt",
  outfile_path = "../tests/data/raw/test_processed.txt",
  data_processing_function = mutate_beta_se_from_zscore_eqtlgen_cis_eqtl,
  chunk_size = 100,
  eqtlgen_cis_eqtl_maf_df = allele_freq
)

# test
test <- fread("../tests/data/raw/test.txt")
test_processed <- fread("../tests/data/raw/test_processed.txt")

test <- mutate_beta_se_from_zscore_eqtlgen_cis_eqtl(test, 
                                                    allele_freq)

if ((!all(test$beta == test_processed$beta)) |
    (!all(test$se == test_processed$se))) {
  stop("Different results produced for beta/se")
}
```

Now with only file paths

```{r}
# write file
append_beta_se_to_eqtlgen_cis_eqtl(
  eqtlgen_cis_eqtl_filepath = "../data/raw/test.txt",
  eqtlgen_cis_eqtl_maf_filepath = "../tests/data/raw/2018-07-18_SNP_AF_for_AlleleB_combined_allele_counts_and_MAF_pos_added.txt",
  outfile_path = "../tests/data/raw/test_processed2.txt",
  chunk_size = 500
)

# test
test_processed2 <- fread("../tests/data/raw/test_processed2.txt")

if ((!all(test$beta == test_processed2$beta)) |
    (!all(test$se == test_processed2$se))) {
  stop("Different results produced for beta/se")
}
```

# Clean up

```{r}
file.remove(c("../tests/data/raw/test.txt",
              "../tests/data/raw/test_processed.txt",
              "../tests/data/raw/test_processed2.txt"))
```

# Process full file

Unhash to run

```{r}
# full file
append_beta_se_to_eqtlgen_cis_eqtl(
  eqtlgen_cis_eqtl_filepath = "../tests/data/raw/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded.txt",
  eqtlgen_cis_eqtl_maf_filepath = "../tests/data/raw/2018-07-18_SNP_AF_for_AlleleB_combined_allele_counts_and_MAF_pos_added.txt",
  outfile_path = "../tests/data/raw/2019-12-11-cis-eQTLsFDR-ProbeLevel-CohortInfoRemoved-BonferroniAdded-beta-se.txt",
  chunk_size = 1000000
)
```

